Data Exploration

### 2019-01-30 src/biological_data_prep.Rmd
moved where to deal with duplicates.

>>>>

# remove any duplicate entries

Check to see how many cells have >1 point per timeslice. This needs to be timesliced and 3D

```{r cell_id_3d}
data_aea$cell_id_3d <- paste(data_aea$cell_id, data_aea$depthlayerno, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(data_aea)
```

and now do one with a timeslice label
```{r cell_id_xyzt}
data_aea$cell_id_xyzt <- paste(data_aea$cell_id_3d, data_aea$year, data_aea$month, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(data_aea)
write.csv(data_aea, "../output/bio/presence_points_without_envdata_relooped_glbathy_cellid_xyzt.csv", row.names = FALSE)
```

ok lets see how many have more than one obs per cell
```{r unique_pointpercell}
unique_pointpercell <- unique(data_aea$total_cell_obs_xyzt)
unique_pointpercell
```

bugger... ok so lets do a frequency table
```{r unique_pointpercell frequencies}
celobsmore1 <- subset(data_aea, total_cell_obs_xyzt >1)
celobsmore1 #2042 rows
pointpercellcount <- count(celobsmore1, "total_cell_obs_xyzt")
pointpercellcount$no_cells <- pointpercellcount$freq / pointpercellcount$total_cell_obs_xyzt
write.csv(pointpercellcount, file = "../output/bio/pointpercellcount_pres.csv")
pointpercellcount
```

so there are 949 3d timesliced cells that contain more than one occurrence, and 2042 rows. 
- different sampling days
- different coordinates
- risk of duplicate entries (but only if two instituions inputted the same data with slightly different coordinates and different reserch program names - see cell3985_18_2000_7 below as example)

Ideally you only want one observation cell cell_id_xyzt. First see if any are missing depth data

```{r}
fmatch("chl_surface", names(celobsmore1)) #28
fmatch("chl_depth", names(celobsmore1)) #29
fmatch("mlp_surface", names(celobsmore1)) #30
fmatch("o2_surface", names(celobsmore1)) #31
fmatch("o2_depth", names(celobsmore1)) #32
fmatch("salinity_surface", names(celobsmore1)) #33
fmatch("salinity_depth", names(celobsmore1)) #34
fmatch("ssh_surface", names(celobsmore1)) #35
fmatch("temp_surface", names(celobsmore1)) #36
fmatch("temp_depth", names(celobsmore1)) #37
```

then subset any with NA values

```{r celobsmore1_novals}
celobsmore1_novals <-  celobsmore1[!complete.cases(celobsmore1[28:37]), ] #28:37 represent the columns w/ env. data
nrow(celobsmore1_novals) #just to check how many there are - 1095 out of 2042
```

ok and subset with all values

```{r celobsmore1_allval}
celobsmore1_allvals <- celobsmore1[complete.cases(celobsmore1[28:37]), ] #28:37 represent the columns w/ env. data
nrow(celobsmore1_allvals) #947 out of 2042
```

ok so how many unique cells do we have with complete data...

```{r}
pointpercellcount_alldata <- count(celobsmore1_allvals, "total_cell_obs_xyzt")
pointpercellcount_alldata$no_cells <- pointpercellcount_alldata$freq / pointpercellcount_alldata$total_cell_obs_xyzt
write.csv(pointpercellcount_alldata, file = "../output/bio/pointpercellcount_alldata.csv")
pointpercellcount_alldata
```

and see which cell_id_xyzt appear in both celobsmore1_allvals and celobsmore1_novals
```{r}
celobsmore1_allnovals <- inner_join(celobsmore1_allvals, celobsmore1_novals)
celobsmore1_allnovals 
```
None.... which is good. I'd expect observations in the same xyzt to have the same values

so, I can successfully reduce the observations down to one per xyzt

```{r}
fmatch("cell_id_xyzt", names(celobsmore1)) #50
duplicatesbegone <- celobsmore1[!duplicated(celobsmore1[ ,50]), ] #creates a df with all duplicates gone (now 1 obs per cell)
nrow(duplicatesbegone) #949 - good
presabnodup <- subset(data_aea, total_cell_obs_xyzt == "1") #remove all rows that have more than one obs
nrow(presabnodup) #9351 - this adds up 
data_aea <- rbind(presabnodup, duplicatesbegone)
nrow(data_aea) #10300 - great!
```

#recalculate the total_cell_obs_xyzt column
```{r}
obs_cell_yymm_depth <- count(data_aea, c("cell_id", "year", "month", "depthlayerno"))
data_aea <- merge(data_aea, obs_cell_yymm_depth, by.x=c("cell_id", "year", "month", "depthlayerno"), by.y=c("cell_id", "year", "month", "depthlayerno"))
names(data_aea)[names(data_aea)=="freq"] <- "total_cell_obs_xyzt"
head(data_aea)
```

check for unique values(should be 1 only)
```{r}
unique_pointpercell <- unique(data_aea$total_cell_obs_xyzt)
unique_pointpercell
```
Great! Save

```{r}
write.csv(data_aea, file = "../output/bio/presence_points_without_envdata_relooped_glbathy_nodup.csv", row.names = FALSE) # change the path to where ever you would like to save it
```

### 2019-01-30 src/data_exploration.Rmd
moved where to deal with duplicates.

# presence points
Read in the presence points

```{r}
presence <- read.csv("../output/bio/presence_points_without_envdata_relooped_glbathy.csv", header = TRUE)
head(presence)
```

@@

# background points
Read in the presence points

```{r}
background <- read.csv("../output/bio/background_complete_obs_cels_globot.csv", header = TRUE)
head(background)
```

@@

first see which nafo zones were sampled in each year
```{r presence nafo by year}
nafo_by_yr <- with(presence, table(year, nafo_zone))
write.csv(nafo_by_yr, file = "../output/bio/nafozone_by_yr.csv")
nafo_by_yr
```

@@

Interesting that there is a point in 1C - this is outside Canadian waters.... anyway
```{r}
onec <-  subset(presence, nafo_zone == "1C")
onec
```

@@

```{r presence by nafo}
nafo0a <- subset(presence, nafo_zone == "0A")
nafo0b <- subset(presence, nafo_zone == "0B")
nafo1c <- subset(presence, nafo_zone == "1C")
nafo2g <- subset(presence, nafo_zone == "2G")
nafo2h <- subset(presence, nafo_zone == "2H")
nafo2j <- subset(presence, nafo_zone == "2J")
nafo3k <- subset(presence, nafo_zone == "3K")
nafo3l <- subset(presence, nafo_zone == "3L")
nafo3m <- subset(presence, nafo_zone == "3M")
nafo3n <- subset(presence, nafo_zone == "3N")
nafo3o <- subset(presence, nafo_zone == "3O")
nafo3ps <- subset(presence, nafo_zone == "3Ps")
nafo4r <- subset(presence, nafo_zone == "4R")
nafo4s <- subset(presence, nafo_zone == "4S")
nafo4t <- subset(presence, nafo_zone == "4T")
nafo4vn <- subset(presence, nafo_zone == "4Vn")
nafo4vs <- subset(presence, nafo_zone == "4Vs")
nafo4w <- subset(presence, nafo_zone == "4W")
nafo4x <- subset(presence, nafo_zone == "4X")
nafo5y <- subset(presence, nafo_zone == "5Y")
nafo5ze <- subset(presence, nafo_zone == "5ZE")
nafohudson <- subset(presence, nafo_zone == "HudsonStrait")
```

plot by each variable, less 3m (2 samples) 1c (one sample) and 5ze (zero samples?!)
```{r presence by nafo by variable}
ggplot(nafo0a, aes(x = temp_surface, colour = nafo_zone)) + geom_density(na.rm = TRUE) + geom_density(data=nafo0b, na.rm = TRUE) + geom_density(data=nafo2g, na.rm = TRUE) + geom_density(data=nafo2h, na.rm = TRUE) + geom_density(data=nafo2j, na.rm = TRUE) + geom_density(data=nafo3k, na.rm = TRUE) + geom_density(data=nafo3l, na.rm = TRUE) + geom_density(data=nafo3n, na.rm = TRUE) + geom_density(data=nafo3o, na.rm = TRUE) + geom_density(data=nafo3ps, na.rm = TRUE) + geom_density(data=nafo4r, na.rm = TRUE) + geom_density(data=nafo4s, na.rm = TRUE) + geom_density(data=nafo4t, na.rm = TRUE) + geom_density(data=nafo4vn, na.rm = TRUE) + geom_density(data=nafo4vs, na.rm = TRUE) + geom_density(data=nafo4w, na.rm = TRUE) + geom_density(data=nafo4x, na.rm = TRUE) + geom_density(data=nafo5y, na.rm = TRUE) + geom_density(data=nafohudson, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_nafo_plots/temp_surface_nafo.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

@@

first a table of how many observations each instituioncode has
```{r institution code analysis - count}
obs_by_ins <- count(presence, "institutioncode")
obs_by_ins
write.csv(obs_by_ins, file = "../output/bio/samplinginstitutions/no_observations_institutioncode.csv")
```
ok so IMR, NAFO, ICES, & USM have very few entries (1, 1, 1, and 3 respectively)

@@

for this you need the joined dataset. this was created under mergig_datasets.src and the file is ../output/bio/presab.csv

```{r}
presab <- read.csv("../output/bio/presab.csv", header = TRUE)
```

@@

vif_allbutnaoamo_sq <- read.csv("../output/bio/vif_allbutnaoamo.csv", header = TRUE)
vif_allbutnaoamo_sq$GVIF2Dfsq <- vif_allbutnaoamo_sq$GVIF..1..2.Df..^2
write.csv(vif_allbutnaoamo_sq, "../output/bio/vif_allbutnaoamo_sq.csv", row.names = TRUE)
vif_allbutnaoamo_sq
```

# check to see how many cells have >1 point per timeslice

this needs to be 3D so you need to create a cell_id_3d
```{r cell_id_3d}
presab$cell_id_3d <- paste(presab$cell_id, presab$depthlayerno, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(presab)
```
and now do one with a timeslice label
```{r cell_id_xyzt}
presab$cell_id_xyzt <- paste(presab$cell_id_3d, presab$year, presab$month, sep="_") #create a new column that is a unique cell_id & depth ID. 
head(presab)
write.csv(presab, "../output/bio/presab_cellid_xyzt.csv", row.names = FALSE)
```

because background points can land in the same xyzt, subset the dataset to be only presences

```{r}
presencedup <- subset(presab, occurrence == "1")
nrow(presencedup)
```

the data is in presencedup$total_cell_obs_xyzt
```{r unique_pointpercell}
unique_pointpercell <- unique(presencedup$total_cell_obs_xyzt)
unique_pointpercell
```
bugger... ok so lets do a frequency table
```{r unique_pointpercell frequencies}
pointpercellcount <- count(celobsmore1, "total_cell_obs_xyzt")
pointpercellcount$no_cells <- pointpercellcount$freq / pointpercellcount$total_cell_obs_xyzt
write.csv(pointpercellcount, file = "../output/bio/pointpercellcount.csv")
pointpercellcount
```

so there are 949 3d timesliced cells that contain more than one occurrence, and 2042 rows. 
- different sampling days
- different coordinates
- risk of duplicate entries (but only if two instituions inputted the same data with slightly different coordinates and different reserch program names - see cell3985_18_2000_7 below as example)

```{r}
cell3985_18_2000_7  <- subset(celobsmore1, cell_id_xyzt == "3985_18_2000_7")
cell3985_18_2000_7 
```

Ideally you only want one observation cell cell_id_xyzt. First see if any are missing depth data

find the column numbers that correspond to the env. data (use fastmatch package as is quicker)

```{r}
fmatch("chl_surface", names(celobsmore1)) #28
fmatch("chl_depth", names(celobsmore1)) #29
fmatch("mlp_surface", names(celobsmore1)) #30
fmatch("o2_surface", names(celobsmore1)) #31
fmatch("o2_depth", names(celobsmore1)) #32
fmatch("salinity_surface", names(celobsmore1)) #33
fmatch("salinity_depth", names(celobsmore1)) #34
fmatch("ssh_surface", names(celobsmore1)) #35
fmatch("temp_surface", names(celobsmore1)) #36
fmatch("temp_depth", names(celobsmore1)) #37
```

then subset any with NA values

```{r celobsmore1_novals}
celobsmore1_novals <-  celobsmore1[!complete.cases(celobsmore1[28:37]), ] #28:37 represent the columns w/ env. data
nrow(celobsmore1_novals) #just to check how many there are - 1095 out of 2042
```

ok and subset with all values

```{r celobsmore1_allval}
celobsmore1_allvals <- celobsmore1[complete.cases(celobsmore1[28:37]), ] #28:37 represent the columns w/ env. data
nrow(celobsmore1_allvals) #947 out of 2042
```

ok so how many unique cells do we have with complete data...

```{r}
pointpercellcount_alldata <- count(celobsmore1_allvals, "total_cell_obs_xyzt")
pointpercellcount_alldata$no_cells <- pointpercellcount_alldata$freq / pointpercellcount_alldata$total_cell_obs_xyzt
write.csv(pointpercellcount_alldata, file = "../output/bio/pointpercellcount_alldata.csv")
pointpercellcount_alldata
```

and see which cell_id_xyzt appear in both celobsmore1_allvals and celobsmore1_novals
```{r}
celobsmore1_allnovals <- inner_join(celobsmore1_allvals, celobsmore1_novals)
celobsmore1_allnovals 
```
None.... which is good. I'd expect observations in the same xyzt to have the same values

so, I can successfully reduce the observations down to one per xyzt

```{r}
fmatch("cell_id_xyzt", names(celobsmore1)) #46
duplicatesbegone <- celobsmore1[!duplicated(celobsmore1[ ,46]), ] #creates a df with all duplicates gone (now 1 obs per cell)
nrow(duplicatesbegone) #949 - good
presabnodup <- subset(presab, total_cell_obs_xyzt == "1") #remove all rows that have more than one obs
nrow(presabnodup) #1549351 - this adds up 
presab <- rbind(presabnodup, duplicatesbegone)
nrow(presab) #1550300 - great!

#recalculate the total_cell_obs_xyzt column

names(presab)[names(presab)=="total_cell_obs_xyzt"] <- "XXtotal_cell_obs_xyzt" #rename col
obs_cell_yymm_depth_dup <- count(presab, cell_id, year, month, depthlayerno)
test <- merge(presab, obs_cell_yymm_depth_dup, by.x=c("cell_id", "year", "month", "depthlayerno"), by.y=c("cell_id", "year", "month", "depthlayerno"))
names(test)[names(test)=="n"] <- "total_cell_obs_xyzt" #rename col

write.csv(presab, "../output/bio/presab_cellid_xyzt_nodup.csv")
rm(presabnodup, duplicatesbegone, celobsmore1, celobsmore1_novals, celobsmore1_allnovals, celobsmore1_allvals) #just because they are bigish variables
```

@@

>>>>

# read in all data

and subset into background and presences
```{r}
presab <- read.csv("../output/bio/presab.csv", header = TRUE)
presence <- subset(presab, occurrence == "1")
background <- subset(presab, occurrence == "0")
```

@@

# background points

@@

first see which nafo zones were sampled in each year
```{r presence nafo by year}
presence$nafo_zone <- as.character(presence$nafo_zone)
nafo_by_yr <- with(presence, table(year, nafo_zone))
write.csv(nafo_by_yr, file = "../output/bio/nafozone_by_yr.csv")
nafo_by_yr
```

@@

Interesting that there is a point in 1C - this is outside Canadian waters remove from the dataset
```{r}
presab <- presab[!(presab$nafo_zone == "1C" & presab$occurrence == "1"), ]
write.csv(presab, "../output/bio/presab.csv", row.names = FALSE)
presence <- presence[!(presence$nafo_zone == "1C"), ]
```

and by month again
```{r prsence nafo by month2}
nafo_by_mth <- with(presence, table(nafo_zone, month))
write.csv(nafo_by_mth, file = "../output/bio/nafozone_by_mth.csv")
nafo_by_mth
```

@@

```{r presence by nafo}
nafo0a <- subset(presence, nafo_zone == "0A")
nafo0b <- subset(presence, nafo_zone == "0B")
nafo2g <- subset(presence, nafo_zone == "2G")
nafo2h <- subset(presence, nafo_zone == "2H")
nafo2j <- subset(presence, nafo_zone == "2J")
nafo3k <- subset(presence, nafo_zone == "3K")
nafo3l <- subset(presence, nafo_zone == "3L")
nafo3m <- subset(presence, nafo_zone == "3M")
nafo3n <- subset(presence, nafo_zone == "3N")
nafo3o <- subset(presence, nafo_zone == "3O")
nafo3ps <- subset(presence, nafo_zone == "3Ps")
nafo4r <- subset(presence, nafo_zone == "4R")
nafo4s <- subset(presence, nafo_zone == "4S")
nafo4t <- subset(presence, nafo_zone == "4T")
nafo4vn <- subset(presence, nafo_zone == "4Vn")
nafo4vs <- subset(presence, nafo_zone == "4Vs")
nafo4w <- subset(presence, nafo_zone == "4W")
nafo4x <- subset(presence, nafo_zone == "4X")
nafo5y <- subset(presence, nafo_zone == "5Y")
nafo5ze <- subset(presence, nafo_zone == "5ZE")
nafohudson <- subset(presence, nafo_zone == "HudsonStrait")
```

plot by each variable, less 3m (2 samples) and 5ze (2 samples)
```{r presence by nafo by variable}
ggplot(nafo0a, aes(x = temp_surface, colour = nafo_zone)) + geom_density(na.rm = TRUE) + geom_density(data=nafo0b, na.rm = TRUE) + geom_density(data=nafo2g, na.rm = TRUE) + geom_density(data=nafo2h, na.rm = TRUE) + geom_density(data=nafo2j, na.rm = TRUE) + geom_density(data=nafo3k, na.rm = TRUE) + geom_density(data=nafo3l, na.rm = TRUE) + geom_density(data=nafo3n, na.rm = TRUE) + geom_density(data=nafo3o, na.rm = TRUE) + geom_density(data=nafo3ps, na.rm = TRUE) + geom_density(data=nafo4r, na.rm = TRUE) + geom_density(data=nafo4s, na.rm = TRUE) + geom_density(data=nafo4t, na.rm = TRUE) + geom_density(data=nafo4vn, na.rm = TRUE) + geom_density(data=nafo4vs, na.rm = TRUE) + geom_density(data=nafo4w, na.rm = TRUE) + geom_density(data=nafo4x, na.rm = TRUE) + geom_density(data=nafo5y, na.rm = TRUE) + geom_density(data=nafohudson, na.rm = TRUE)
dev.copy(png,"../output/env/env_by_nafo_plots/temp_surface_nafo.png") # to automatically save the plot to a png AND show it inline
dev.off() # stops automatic saving of the plot to a png

@@

first a table of how many observations each instituioncode has
```{r institution code analysis - count}
obs_by_ins <- count(presence, "institutioncode")
obs_by_ins
write.csv(obs_by_ins, file = "../output/bio/samplinginstitutions/no_observations_institutioncode.csv")
```
ok so NEFSC and ROM only have one point each

@@

for this you need the joined dataset.

@@

vif_allbutnaoamo_sq <- read.csv("../output/bio/vif_allbutnaoamo.csv", header = TRUE)
vif_allbutnaoamo_sq$GVIF2Dfsq <- vif_allbutnaoamo_sq$GVIF..1..2.Df..^2
write.csv(vif_allbutnaoamo_sq, "../output/bio/vif_allbutnaoamo_sq.csv", row.names = TRUE)
vif_allbutnaoamo_sq
```
not much diff..

@@

Run a VIF for each month (with and without gear)
